---
title: "CuratedAtlasQueryR"
output: "html_vignette"
vignette: >
  %\VignetteIndexEntry{CuratedAtlasQueryR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
knit: >
    (function(x, ...){
        proj_root <- rprojroot::find_package_root_file() |> normalizePath()
        rmarkdown::render(
            x,
            output_format = "html_document"
        )
        rmarkdown::render(
            x,
            output_file = "README.md",
            output_format = "github_document",
            output_dir = proj_root,
            knit_root_dir = proj_root
        )
    })
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
root_dir <- knitr::opts_knit$get("root.dir")
if (!is.null(root_dir)){
    # This hack fixes the relative image paths. 
    # See https://github.com/rstudio/rmarkdown/issues/2473
    knitr::opts_knit$set(
        output.dir = root_dir
    )
}
proj_root <- rprojroot::find_package_root_file() |> normalizePath() 
# Utility function for figures to force them to have the correct path
find_figure <- function(names){
    rprojroot::find_package_root_file() |>
        file.path("man", "figures", names) 
}
```

<!-- badges: start -->
[![Lifecycle:maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
<!-- badges: end -->

`CuratedAtlasQuery` is a query interface that allow the programmatic exploration and retrieval of the harmonised, curated and reannotated CELLxGENE single-cell human cell atlas. Data can be retrieved at cell, sample, or dataset levels based on filtering criteria. 


```{r, echo=FALSE, out.height = c("139px"), out.width = "120x" }
find_figure("logo.png") |> knitr::include_graphics()
```

```{r, echo=FALSE, out.height = c("58px"), out.width = c("155x", "129px", "202px", "219px", "180px")}
c(
  "svcf_logo.jpeg", 
  "czi_logo.png", 
  "bioconductor_logo.jpg",
  "vca_logo.png",
  "nectar_logo.png"
) |> 
    find_figure() |>
    knitr::include_graphics()
```

# Query interface

## Installation

```{r, eval=FALSE}
devtools::install_github("stemangiola/CuratedAtlasQueryR")
```

## Load the package

```{r, message=FALSE, warning=FALSE}
library(CuratedAtlasQueryR)
```

## Load and explore the metadata

### Load the metadata

```{r}
metadata <- get_metadata()
metadata
```

The `metadata` variable can then be re-used for all subsequent queries.

### Explore the tissue 

```{r, eval=FALSE}
metadata |>
    dplyr::distinct(tissue, file_id) 
```

```{r}
#> # Source:     SQL [?? x 2]
#> # Database:   sqlite 3.40.0 [public_access@zki3lfhznsa.db.cloud.edu.au:5432/metadata]
#> # Ordered by: desc(n)
#>    tissue                      n
#>    <chr>                 <int64>
#>  1 blood                      47
#>  2 heart left ventricle       46
#>  3 cortex of kidney           31
#>  4 renal medulla              29
#>  5 lung                       27
#>  6 liver                      24
#>  7 middle temporal gyrus      24
#>  8 kidney                     19
#>  9 intestine                  18
#> 10 thymus                     17
#> # … with more rows
```


## Download single-cell RNA sequencing counts 

### Query raw counts

```{r}
single_cell_counts = 
    metadata |>
    dplyr::filter(
        ethnicity == "African" &
        stringr::str_like(assay, "%10x%") &
        tissue == "lung parenchyma" &
        stringr::str_like(cell_type, "%CD4%")
    ) |>
    get_SingleCellExperiment()

single_cell_counts
```

### Query counts scaled per million

This is helpful if just few genes are of interest, as they can be compared across samples.

```{r}
single_cell_counts = 
    metadata |>
    dplyr::filter(
        ethnicity == "African" &
        stringr::str_like(assay, "%10x%") &
        tissue == "lung parenchyma" &
        stringr::str_like(cell_type, "%CD4%")
    ) |>
    get_SingleCellExperiment(assays = "cpm")

single_cell_counts
```

### Extract only a subset of genes

```{r}
single_cell_counts = 
    metadata |>
    dplyr::filter(
        ethnicity == "African" &
        stringr::str_like(assay, "%10x%") &
        tissue == "lung parenchyma" &
        stringr::str_like(cell_type, "%CD4%")
    ) |>
    get_SingleCellExperiment(assays = "cpm", features = "PUM1")

single_cell_counts
```

### Extract the counts as a Seurat object

This convert the H5 SingleCellExperiment to Seurat so it might take long time and occupy a lot of memory depending on how many cells you are requesting.


```{r}
single_cell_counts = 
    metadata |>
    dplyr::filter(
        ethnicity == "African" &
        stringr::str_like(assay, "%10x%") &
        tissue == "lung parenchyma" &
        stringr::str_like(cell_type, "%CD4%")
    ) |>
    get_seurat()

single_cell_counts
```

## Save your `SingleCellExperiment`

The returned `SingleCellExperiment` can be saved with two modalities, as `.rds` or as `HDF5`.

### Saving as RDS (fast saving, slow reading)

Saving as `.rds` has the advantage that is very fast, the `.rds` file occupies very little disk space as it only stored the links for the files in your ache.

However it has the disadvantage that for big `SingleCellExperiment` objects, which merge a lot of HDF5 from your `get_SingleCellExperiment` the display and manipulation is going to be slow.

```{r, eval=FALSE}
single_cell_counts |> saveRDS("single_cell_counts.rds")
```

### Saving as HDF5 (slow saving, fast reading)

Saving as `.rds` has the advantage that rewrites on disk a monolithic `HDF5` and so displaying and manipulating large `SingleCellExperiment` objects, which merge a lot of HDF5 from your `get_SingleCellExperiment`, is going to be fast.

However it has the disadvantage that the files are going to be larger as they include the count information, and the saving process is going to be slow for large objects.

```{r, eval=FALSE}
single_cell_counts |> saveHDF5SummarizedExperiment("single_cell_counts")
```

## Visualise gene transcription

We can gather all CD14 monocytes cells and plot the distribution of HLA-A across all tissues

```{r, eval=FALSE, echo=FALSE}
library(tidySingleCellExperiment)
library(ggplot2)

# Plots with styling

# Plot by disease
metadata |>
  # Filter and subset
  filter(cell_type_harmonised=="cd14 mono") |>
  filter(file_id_db != "c5a05f23f9784a3be3bfa651198a48eb") |> 
  
  # Get counts per million for HCA-A gene
  get_SingleCellExperiment(assays = "cpm", features = "HLA-A") |> 
  
  # Add feature to table
  join_features("HLA-A", shape = "wide") |> 
    
  # Rank x axis
  as_tibble() |> 
  with_groups(disease, ~ .x |> mutate(median_count = median(`HLA.A`, rm.na=TRUE))) |> 
  
  # Plot
  ggplot(aes( fct_reorder(disease, median_count,.desc = TRUE), `HLA.A`,color = file_id)) +
  geom_jitter(shape=".") +
    
  # Style
  guides(color="none") +
  scale_y_log10() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)) + 
  xlab("Disease") + 
  ggtitle("HLA-A in CD14 monocytes by disease") 

# Plot by tissue
metadata |>
  # Filter and subset
  filter(cell_type_harmonised=="cd14 mono") |>
  filter(file_id_db != "c5a05f23f9784a3be3bfa651198a48eb") |> 
  
  # Get counts per million for HCA-A gene
  get_SingleCellExperiment(assays = "cpm", features = "HLA-A") |> 
  
  # Add feature to table
  join_features("HLA-A", shape = "wide") |> 
    
  # Rank x axis
  as_tibble() |> 
  with_groups(tissue_harmonised, ~ .x |> mutate(median_count = median(`HLA.A`, rm.na=TRUE))) |> 
  
  # Plot
  ggplot(aes( fct_reorder(tissue_harmonised, median_count,.desc = TRUE), `HLA.A`,color = file_id)) +
  geom_jitter(shape=".") +
    
  # Style
  guides(color="none") +
  scale_y_log10() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)) + 
  xlab("Tissue") + 
  ggtitle("HLA-A in CD14 monocytes by tissue") 

```

```{r, eval=FALSE}
library(tidySingleCellExperiment)
library(ggplot2)

metadata |>
  # Filter and subset
  filter(cell_type_harmonised=="cd14 mono") |>

  # Get counts per million for HCA-A gene
  get_SingleCellExperiment(assays = "cpm", features = "HLA-A") |> 
  
  # Plot (styling code have been omitted)
  join_features("HLA-A", shape = "wide") |> 
  ggplot(aes( disease, `HLA.A`,color = file_id)) +
  geom_jitter(shape=".") 
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
find_figure("HLA_A_disease_plot.png") |> knitr::include_graphics()
```

```{r, eval=FALSE}

metadata |> 
    
  # Filter and subset
  filter(cell_type_harmonised=="nk") |> 

  # Get counts per million for HCA-A gene 
  get_SingleCellExperiment(assays = "cpm", features = "HLA-A") |> 

    # Plot (styling code have been omitted)
  join_features("HLA-A", shape = "wide") |> 
  ggplot(aes( tissue_harmonised, `HLA.A`,color = file_id)) +
  geom_jitter(shape=".") 

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
find_figure("HLA_A_tissue_plot.png") |> knitr::include_graphics()
```

## Obtain Unharmonised Metadata

Various metadata fields are *not* common between datasets, so it does not
make sense for these to live in the main metadata table. However, we can
obtain it using the `get_unharmonised_metadata()` function. This function
returns a data frame with one row per dataset, including the `unharmonised` 
column which contains unharmnised metadata as a nested data frame.

```{r}
harmonised <- get_metadata() |> dplyr::filter(tissue == "kidney blood vessel")
unharmonised <- get_unharmonised_metadata(harmonised)
unharmonised
```

Notice that the columns differ between each dataset's data frame:

```{r}
dplyr::pull(unharmonised) |> head(2)
```

# Cell metadata

Dataset-specific columns (definitions available at cellxgene.cziscience.com)

`cell_count`, `collection_id`, `created_at.x`, `created_at.y`, `dataset_deployments`, `dataset_id`, `file_id`, `filename`, `filetype`, `is_primary_data.y`, `is_valid`, `linked_genesets`, `mean_genes_per_cell`, `name`, `published`, `published_at`, `revised_at`, `revision`, `s3_uri`, `schema_version`, `tombstone`, `updated_at.x`, `updated_at.y`, `user_submitted`, `x_normalization`

Sample-specific columns (definitions available at cellxgene.cziscience.com)

`sample_`, `sample_name`, `age_days`, `assay`, `assay_ontology_term_id`, `development_stage`, `development_stage_ontology_term_id`, `ethnicity`, `ethnicity_ontology_term_id`, `experiment___`, `organism`, `organism_ontology_term_id`, `sample_placeholder`, `sex`, `sex_ontology_term_id`, `tissue`, `tissue_harmonised`, `tissue_ontology_term_id`, `disease`, `disease_ontology_term_id`, `is_primary_data.x`

Cell-specific columns (definitions available at cellxgene.cziscience.com)

`cell_`, `cell_type`, `cell_type_ontology_term_idm`, `cell_type_harmonised`, `confidence_class`, `cell_annotation_azimuth_l2`, `cell_annotation_blueprint_singler` 

Through harmonisation and curation we introduced custom column, not present in the original CELLxGENE metadata

- `tissue_harmonised`: a coarser tissue name for better filtering
- `age_days`: the number of days corresponding to the age
- `cell_type_harmonised`: the consensus call identity (for immune cells) using the original and three novel annotations using Seurat Azimuth and SingleR
- `confidence_class`: an ordinal class of how confident `cell_type_harmonised` is. 1 is complete consensus, 2 is 3 out of four and so on.             
- `cell_annotation_azimuth_l2`: Azimuth cell annotation
- `cell_annotation_blueprint_singler`: SingleR cell annotation using Blueprint reference
- `cell_annotation_blueprint_monaco`: SingleR cell annotation using Monaco reference
- `sample_id_db`: Sample subdivision for internal use
- `file_id_db`: File subdivision for internal use
- `sample_`: Sample ID
- `.sample_name`: How samples were defined

# RNA abundance

The `raw` assay includes RNA abundance in the positive real scale (not transformed with non-linear functions, e.g. log sqrt). Originally CELLxGENE include a mix of scales and transformations specified in the `x_normalization` column.

The `cpm` assay includes counts per million.
